%-----------------------------------
%
\chapter{Tokenization}
%
%-----------------------------------

\section{Intro}


Tokenization is a step in natural language processing (NLP) that splits text into smaller pieces called \textit{tokens}.
These tokens are the basic units of data processed by language models, and different tokenization strategies have been explored in literature.
For instance, consider the phrase \texttt{"I love NLP"}, which may be split by different tokenization schemes as follows:
\begin{itemize}
    \item \textbf{Word-level:}: \texttt{["I", "love", "NLP"]}

    \item \textbf{Character-level:} \texttt{["I", " ", "l", "o", "v", "e", " ", "N", "L", "P"]}

    \item \textbf{Subword-level:} \texttt{["I", "lo", "ve", "NL", "P"]} or \texttt{["I", "love", "N", "LP"]}

\end{itemize}
In standard NLP pipelines, an embedding function then maps each token into a vector in \(\mathbb{R}^{d}\), where \(d\) is called the \textit{embedding dimension}, \textit{hidden dimension}, or \textit{latent dimension}.
In GPT-2, for instance, one might have \(d = 768\)~\citep{radford2019language}, and in Llama-based models, we may have \(d = 4096\)~\citep{touvron2023llama}.
A simplified but representative text-to-vector pipeline is as follows:
\begin{equation*}
    \underbrace{\texttt{"I love NLP"}}_{\text{input text}}
    \xrightarrow{\text{Tokenizer}}
    \underbrace{\texttt{["I", "love", "NLP"]}}_{\text{tokens}}
    \xrightarrow{\text{Embed}}
    \underbrace{
        \begin{bmatrix}
            0.94   & \cdots & - 1.60 \\
            - 0.57 & \cdots & 2.58   \\
            0.34   & \cdots & 0.05
        \end{bmatrix}
    }_{\text{embedded inputs in \(\mathbb{R}^{3 \times d}\)}}
\end{equation*}

In the above, we have mapped each token to a vector in \(\mathbb{R}^{d}\) and laid them out horizontally in a single matrix following convention, where the numbers were arbitrarily chosen for illustrative purposes.
We emphasize that the above is a simplified presentation of a complex process, and we refer to~\citep{mielke2021between} for a comprehensive survey and history of various tokenization methods in NLP.



For theoretical analyses of transformers, one usually assumes that a vectorized embedding of the text is given, i.e., the input text is already processed by some tokenization method and embedding function.
This is because tokenization and embedding strategies are often model and task-specific, and so are commonly abstracted as a pre-processing step.
Although this simplification is convenient for theoretical analysis, the details of tokenizations and embeddings should be considered in practical implementations.





\section{Tokenization Methods}

\subsection{Byte Pair Encoding (BPE)}

Byte pair encoding~\citep{gage1994new,sennrich2015neural} is a class of methods for data compression or text segmentation that iteratively introduces new symbols by merging the most frequent pairs in a sequence.
It is used to construct the vocabulary in language modeling, where it is applied to the corpus prior to training.
One objective of BPE is to better capture rare words that would otherwise be out-of-vocabulary during deployment, without significantly expanding the overall vocabulary size.
As an example, suppose that our initial corpus is as follows:
\begin{center}
\begin{minipage}{0.5\linewidth}
\begin{verbatim}
aaabdaaabac
\end{verbatim}
\end{minipage}
\end{center}
where we begin with the initial vocabulary of \texttt{["a", "b", "c"]}.
The pair (bigram) \texttt{"aa"} occurs most often, so we replace it with a new token \texttt{"Z"}, which results in:
\begin{center}
\begin{minipage}{0.5\linewidth}
\begin{verbatim}
ZabdZabac, Z=aa
\end{verbatim}
\end{minipage}
\end{center}
We then repeat this by substituting the pair \texttt{"ab"} with \texttt{"Y"}:
\begin{center}
\begin{minipage}{0.5\linewidth}
\begin{verbatim}
ZYdZYac, Y=ab, Z=aa
\end{verbatim}
\end{minipage}
\end{center}
And finally, replacing \texttt{"ZY"} with \texttt{"X"} we get:
\begin{center}
\begin{minipage}{0.5\linewidth}
\begin{verbatim}
XdXac, X=ZY, Y=ab, Z=aa
\end{verbatim}
\end{minipage}
\end{center}
Because all bigrams have the same frequency, no more substitutions may be made, and so we terminate with the final vocabulary of \texttt{["a", "b", "c", "X", "Y", "Z"]}.
The above is a simplified example with an initial vocabulary of single-character tokens, which may more generally also include unicode characters or sequences.
Moreover, one may also define more complex new-symbol and termination conditions, so as to better control the final vocabulary.
Other methods of forming a language model vocabulary include WordPiece~\citep{schuster2012japanese} and unigram tokenization~\citep{kudo2018subword}.




\section{Additional Symbols and Representational Capacity}

While the effect of tokenization schemes on the representational capacity of language models is not well-understood, the \emph{alphabet} that a transformer works over is known to have a significant impact on the model's ability to represent certain models of computation.
A classic example of this is how transformers can simulate Turing machines by using extra symbols to represent their internal configuration \citep{perez-etal-2021-turing}.
Intuitively, in a sequence-to-sequence or a language modeling task, the transformer uses the richer alphabet of internal representations to output the (changes to the) current internal state of the computational model and access those changes as the output string is fed ot the model as input in the future.

Concretely, rather than working over the alphabet of input symbols $\alphabet$, the transformer (deterministically) outputs symbols from an \emph{augmented} alphabet $\augalphabet = \alphabet \times \intalphabet$, where $\intalphabet$ can be thought of as the alphabet whose symbols represent the (changes to the) internal state of the computational model.
Elements of $\augalphabet$ then represent pairs of string symbols and internal states, and the transformer can use the internal state to keep track of the current state of the computational model while still processing the input string.
As a concrete example, \citet{perez-etal-2021-turing} use an internal alphabet $\intalphabet$ that, among other things, contains the current state of the Turing machine, the content written on the tape, and the position of the tape head.
The same principle can also be used to simulate simpler sequential models of computation, such as finite automata.
In this case, the internal alphabet $\intalphabet$ simply contains the current state of the automaton, meaning that the elements $\left(\sym, \fsaState\right) \in \augalphabet$ represent the symbol $\sym$ on the transition taken by the automaton and the state $\fsaState$ transitioned into the automaton.
Simulating the automaton is then particularly simple: The transformer can simply attend to the previously output internal state to determine the current state of the simulated automaton and then process the strings accordingly.
While not discussed by \citet{perez-etal-2021-turing} initially, outputting the internal states from $\intalphabet$ can be thought of as a form of performing \defn{chain-of-thought} reasoning \citep{wei2023chain}, where the transformer can keep track of the internal state of the computational model as it processes the input string \citep{feng2023revealing,merrill2024the,nowak-etal-2024-representational}.


\subsection{BOS}\label{sec:BOS}

The use of a special symbol such as BOS (beginning of sequence), EOS (end of sequence), and CLS (classification) These are special symbols outside of the usual alphabet of symbols. These special symbols are often used as distinguished positions to carry information about the input - for instance, one might take the output at the CLS position as an embedding of the entire input sentence. Having one of these tokens in the sequence also serves the functional purpose of allowing a transformer to create the value $\frac{1}{n}$ (or $\frac{1}{i+1}$ if future-masked) by using uniform attention \cref{sec:uniform-attention}. For instance, constructing $\frac{1}{i+1}$ was used in \citet{merrill-sabharwal-2024-cot} to do retrieval using the layer-norm hash.


\subsection{Notes on Expressivity}

\AY{Will write}